A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-04-15 03-semantics.tex
Cleanup and compress stuff in Section 3 up to Determinism

\def\JOT{1.5\jot}

@@

In this section, we introduce a reduced, abstract syntax for \CEU and
present an operational semantics that formalizes the behavior of its
programs.
% We describe a small synchronous kernel highlighting the peculiarities of \CEU,
% in particular, the stack-based execution for internal events.
%
The semantics deals only with the control aspects of \CEU.
Side-effects and system calls are encapsulated in an abstract memory access
primitive and are assumed to behave like in conventional imperative
languages.

@@

The grammar below defines the syntax of a subset of \CEU that is
sufficient to describe all peculiarities of the language.
\bgroup
\def\lbl#1{\quad\text{\emph{#1}}}%
\newdimen\X
\X=-1.25\jot
\begin{alignat*}{2}
  P\Coloneqq
      &\enspace\ceu{\Mem(\Id)}
      &&\lbl{any memory access to~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\AwaitExt(\Id)}
      &&\lbl{await external event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\AwaitInt(\Id)}
      &&\lbl{await internal event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\EmitInt(\Id)}
      &&\lbl{emit internal event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\Break}
      &&\lbl{loop escape}\\[\X]
      %%
  \mid&\enspace\ceu{\IfElse{\Mem(\Id)}{P_1}{P_2}}
      &&\lbl{conditional}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\,;\,P_2}
      &&\lbl{sequence}\\[\X]
      %%
  \mid&\enspace\ceu{\Loop P_1}
      &&\lbl{repetition}\\[\X]
      %%
  \mid&\enspace\ceu{\Every{\Id}\ P_1}
      &&\lbl{event iteration}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\And P_2}
      &&\lbl{par/and}\\[\X]
      %%
  \mid&\enspace\ceu{P_2\Or P_2}
      &&\lbl{par/or}\\[\X]
      %%
  \mid&\enspace\ceu{\Fin P}
      &&\lbl{finalization}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\AtLoop P_2}
      &&\lbl{unwinded loop}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\AtAnd\ P_2}
      &&\lbl{unwinded par/and}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\AtOr\ P_2}
      &&\lbl{unwinded par/or}\\[\X]
      %%
  \mid&\enspace\ceu{\CanRun(n)}
      &&\lbl{can run on stack level~$n$}\\[\X]
      %%
  \mid&\enspace\ceu{\Nop}
      &&\lbl{terminated program}
\end{alignat*}
\egroup

The~$\ceu{\Mem(id)}$ primitive represents any access, assignment, or system
call that affects the memory location identified by~$id$.\footnotemark\
%
Following the synchronous hypothesis, $\ceu{\Mem}$ expressions are
considered to be atomic and instantaneous.
%
% As the challenging parts of \CEU reside on its control structures, we are not
% concerned here with a precise semantics for side effects, but only with their
% occurrences in programs.
%
%The special notation $nop$ is used to represent an innocuous $mem$ expression
%(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an
%unused identifier).
%
\footnotetext{Although the same symbol~$\ceu{\Id}$ is used in their
  definition, the $\ceu{\Mem}$, $\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$
  and~$\ceu{\EmitInt}$ expressions do not share identifiers: any identifier
  is either a variable, an external event, or an internal event.}

Most expressions in the abstract language are mapped to their counterparts
in the concrete language.  The exceptions are the finalization
block~$\ceu{\Fin{p}}$ and the \texttt{@}-expressions which do not exist in
the concrete language.  (These expressions result from the expansion of the
transition rules to be presented.)

A further difference between the concrete and abstract languages regards the
emit-await pair.  In the concrete language the \code{await} and \code{emit}
expressions support communication of values between them, e.g., an
``\code{emit a(10)}'' awakes a ``\code{v=await a}'' setting
variable~\code{v} to~10.  Although the abstract $\ceu{\AwaitInt}$
and~$\ceu{\EmitInt}$ expression do not support such communication of values,
it can be easily simulated: we can use a shared variable to hold the value
of an $\ceu{\EmitInt}$ and access it after the corresponding
$\ceu{\AwaitInt}$ awakes.

Finally, a ``\code{finalize $A$ with $B$ end; $C$}'' in the concrete
language is equivalent to ``\ceu{A;\;((\Fin{B})\ \Or\ C)}'' in the abstract
language.  In the concrete language, $A$ and~$C$ execute in sequence, and
the finalization code~$B$ is implicitly suspended waiting for~$C$
termination.  In the abstract language, ``$\ceu{\Fin B}$'' suspends forever
when reached (it is an awaiting expression that never awakes).  Hence, we
need an explicit \code{or} to execute~$C$ in parallel, whose termination
aborts ``$\ceu{\Fin B}$'', which finally causes~$B$ to execute (by the
semantic rules below).

@@

The core of our semantics describes how a program reacts to a single external
input event, i.e., starting from an input event, how the program behaves and
becomes idle again to proceed to a subsequent reaction.
%
We use a set of small-step operational rules~\cite{Plotkin-G-D-1981} which
ensure that at most one transition is possible at any time, resulting in
deterministic reactions.
%
The transition rules map a triple with a program~$p$, a stack level~$n$, and an
emitted event~$e$ to a modified triple as follows:
\[
  \<p,n,e>\trans\<p',n',e'>\,,
\]
where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are non-negative
integers representing the current stack level, and~$e,e'\in\E\cup\{\nil\}$ are
the events emitted before and after the transition (both possibly being the
empty event~$\nil$).
%$\E$ is a set of primitive events.

We will refer to the triples on the left-hand and right-hand sides of
symbol~$\trans$ as \emph{descriptions} (denoted~$\delta$).  The description
on the left-hand side of~$\trans$ is the \emph{input description},
and the one on the right-hand side is the \emph{output description}.

@@

At the beginning of a reaction to an input event~$id$, the input description is
initialized with stack level~0 ($n=0$) and with the externally emitted event~$e=id$.
%, but \code{emitInt} expressions can increase the stack level.
At the end of a reaction, after an arbitrary but finite number of transitions,
the last output description will block with a (possibly) modified program~$p'$ at stack
level~0 and with no event emitted:
\[
  \<p,0,e>\mathbin{\trans[*]}\<p',0,\nil>\,.
\]

We distinguish between two types of transition rules:
    \emph{outermost transitions} $\out$ and
    \emph{nested transitions} $\nst$\,.

@@

The~$\out$ rules \R{push} and \R{pop} are non-recursive definitions that
apply to the program as a whole.  These are the only rules that manipulate
the stack level.
\begin{align*}
  &\AxiomC{$e\ne\nil$}
  \UnaryInfC{$\<p,n,e>\out\<\bcast(p,e),n+1,\nil>$}
  \DisplayProof
  \Rtag{push}\\[\JOT]
  %%
  &\AxiomC{$n>0$}
  \AxiomC{$p=\ceu{\Nop} \vee p=\ceu{\Break} \vee \isblocked(p,n)$}
  \BinaryInfC{$\<p,n,\nil>\out\<p,n-1,\nil>$}
  \DisplayProof
  \Rtag{pop}
\end{align*}

@@

Rule \R{push} matches whenever there is an emitted event in the input
description,
and instantly broadcasts the event to the program, which means
    (a)~awaking active $\ceu{\AwaitExt}$ or $\ceu{\AwaitInt}$ expressions altogether (see function~$\bcast$ in
        Figure~\ref{fig.bcast}),
    (b)~creating a nested reaction by increasing the stack level, and, at the same time,
    (c)~consuming the event ($e$ becomes~$\nil$).
%
Rule \R{push} is the only rule in the semantics that matches an
emitted event and also immediately consumes it.

Rule \R{pop} only decreases the stack level, not affecting the
program, and only applies if the program is blocked (see function~$\isblocked$
in Figure~\ref{fig.isblocked}) or terminated ($p=\ceu{\Nop}$ or $p=\ceu{\Break}$).
This condition ensures that an $\ceu{\EmitInt}$ only resumes after its internal
reaction completes and blocks in the current stack level.

At the beginning of a reaction, an external event is emitted, which
will trigger rule \R{push}, which will immediately raise the stack level
to~1.
At the end of the reaction, the program will block or terminate and
successive applications of
rule~\R{pop} will eventually lead to a description containing this
same program at stack level~0.

@@

A complete reaction consists of a series of transitions of the form
\begin{align*}
  \<p,0,e_\ext>\outpush\<p_1,1,\nil>
  \Big[\null\nst[*]\null\out\null\Big]\!\!\ast
  \null\nst[*]\null\outpop\<p',0,\nil>\,.
\end{align*}
%
%-
% \begin{align*}
% a) &\5\5
%     \LL p,0,ext \RR
%         \1\xrightarrow[out]{push}\1
%     \LL q,1,\epsilon \RR
% \\
% b) &\5\5 \1[ \1\xrightarrow[in]{*}\1
%     \LL r,i,e \RR
%         \1\xrightarrow[out]\1
%     \LL s,j,\epsilon \RR \1]*
% \\
% c) &\5\5 \1\xrightarrow[in]{*}\1
%     \LL t,k,\epsilon \RR
%         \1\xrightarrow[out]{pop}\1
%     \LL u,0,\epsilon \RR
% \end{align*}
%-
%
First, a~$\outpush$ starts a nested reaction at level~1.
Then, a series of alternations between zero or more~$\nst$ transitions (nested reactions) and a
single~$\out$ transition (stack operation) takes place.
Finally, a last~$\outpop$ transition decrements the
stack level to~0 and terminates the reaction.

@@

A $\ceu{\Mem}$ operation becomes a $\ceu{\Nop}$ which indicates the memory
access (rule \R{mem}).
An $\ceu{\EmitInt(id)}$ generates an event $\ceu{\Id}$ and transits to a
$\ceu{\CanRun(n)}$ which can only resume at level~$n$ (rule \R{emit-int}).
Since all~$\nst$ rules can only transit with $e=\nil$, an $\ceu{\EmitInt}$
inevitably causes rule \R{push} to execute at the outer level, creating a new
level~$n+1$ on the stack.
Also, with the new stack level, the resulting $\ceu{\CanRun}(n)$ itself cannot
transit yet (rule~\R{can-run}), providing the desired stack-based semantics for
internal events.

The rules for conditionals and sequences are straightforward:
%
\begin{gather*}
  \AxiomC{$\eval(\ceu{\Mem(\Id)})$}
  \UnaryInfC{$\<\ceu{\IfElse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<p,n,\nil>$}
  \DisplayProof
  \Rtag{if-true}\\[\JOT]
  %%
  \AxiomC{$\lnot\eval(\ceu{\Mem(\Id)})$}
  \UnaryInfC{$\<\ceu{\IfElse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<q,n,\nil>$}
  \DisplayProof
  \Rtag{if-false}\\[\JOT]
  %%
  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
  \UnaryInfC{$\<\ceu{p\,;\,q},n,\nil>\nst\<\ceu{p';\,q},n,e>$}
  \DisplayProof
  \Rtag{seq-adv}
\end{gather*}
\vskip-\belowdisplayskip
\vskip-\abovedisplayskip
\vskip\JOT
\begin{align*}
  \<\ceu{\Nop;\,q},n,\nil>&\nst\<q,n,\nil>\Rtag{seq-nop}\\[\JOT]
  %%
  \<\ceu{\Break;\,q},n,\nil>&\nst\<\ceu{\Break},n,\nil>\Rtag{seq-brk}
\end{align*}

@@

Given that our semantics focuses on control, rules \R{if-true} and
\R{if-false} are the only to query~$\ceu{\Mem}$ expressions.
%
Function~$\eval$ evaluates a given~$\ceu{\Mem}$ expression to a boolean value.
%
%Although the value here is arbitrary, it is unique in a reaction, because a
%given expression can execute only once within it (remember that $loops$ must
%contain $awaits$ which, from rule \textbf{await}, cannot awake in the same
%reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).

@@

The rules for~$\ceu{\And}$ and~$\ceu{\Or}$ compositions ensure that the
left-hand side of compositions always transition before their right-hand
side:
%%
\begin{gather*}
  \hskip-.9em
  \<\ceu{p\And{q}},n,\nil>
  \nst\<\ceu{p\AtAnd(\CanRun(n);q)},n,\nil>
  \Rtag{and-expd}\\[\JOT]
  %%
  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
  \UnaryInfC{$\<\ceu{p\AtAnd{q}},n,\nil>\nst\<\ceu{p'\AtAnd{q}},n,e>$}
  \DisplayProof
  \Rtag{and-adv1}\\[\JOT]
  %%
  \AxiomC{$\isblocked(p,n)$}
  \AxiomC{$\<q,n,\nil>\nst\<q',n,e>$}
  \BinaryInfC{$\<\ceu{p\AtAnd{q}},n,\nil>\nst\<\ceu{p\AtAnd{q'}},n,e>$}
  \DisplayProof
  \Rtag{and-adv2}
\end{gather*}

\begin{gather*}
  \<\ceu{p\Or{q}},n,\nil>
  \nst\<\ceu{p\AtOr(\CanRun(n);q)},n,\nil>
  \Rtag{or-expd}\\[\JOT]
  %%
  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
  \UnaryInfC{$\<\ceu{p\AtOr{q}},n,\nil>\nst\<\ceu{p'\AtOr{q}},n,e>$}
  \DisplayProof
  \Rtag{or-adv1}\\[\JOT]
  %%
  \AxiomC{$\isblocked(p,n)$}
  \AxiomC{$\<q,n,\nil>\nst\<q',n,e>$}
  \BinaryInfC{$\<\ceu{p\AtOr{q}},n,\nil>\nst\<\ceu{p\AtOr{q'}},n,e>$}
  \DisplayProof
  \Rtag{or-adv2}
\end{gather*}

@@

Rules~\R{and-expd} and~\R{or-expd} insert a~$\ceu{\CanRun(n)}$ at the beginning
of the right branch.
This ensures that~any $\ceu{\EmitInt}$ on the left branch, which transits to
a~$\ceu{\CanRun(n)}$, still resumes before the right branch starts.
%
The deterministic behavior of the semantics relies on the \emph{isblocked}
predicate (see Figure~\ref{fig.isblocked}) which appears in rules
\R{and-adv2} and \R{or-adv2}.
These rules require the left branch~$p$ to be blocked for the
right branch to transition from~$q$ to~$q'$.

In a parallel~$\ceu{\AtAnd}$, if one of the sides terminates, the composition is
simply substituted by the other side (rules \R{and-nop1} and
\R{and-nop2} below).
%
In a parallel~$\ceu{\AtOr}$, however, if one of the sides terminates, the
whole composition
terminates and function~$\clear$ is used to properly finalize the aborted
side (rules \R{or-nop1} and \R{or-nop2}).
\begin{gather*}
  \<\ceu{{\Nop}\AtAnd{q}},n,\nil>\nst\<q,n,\nil>\Rtag{and-nop1}\\[\JOT]
  %%
  \AxiomC{$\isblocked(p,n)$}
  \UnaryInfC{$\<\ceu{p\AtAnd{\Nop}},n,\nil>\nst\<p,n,\nil>$}
  \DisplayProof
  \Rtag{and-nop2}\\[\JOT]
  %%
  \<\ceu{{\Nop}\AtOr{q}},n,\nil>\nst\<\clear(q),n,\nil>\Rtag{or-nop1}\\[\JOT]
  %%
  \AxiomC{$\isblocked(p,n)$}
  \UnaryInfC{$\<\ceu{p\AtOr{\Nop}},n,\nil>\nst\<\clear(p),n,\nil>$}
  \DisplayProof
  \Rtag{or-nop2}
\end{gather*}

@@

The~$\clear$ function (see Figure~\ref{fig.clear}) concatenates all
active~$\ceu{\Fin}$ bodies of the side being aborted, so that they execute before the
composition rejoins.
Note that there are no transition rules for~$\ceu{\Fin}$ expressions.
This is because once reached, a $\ceu{\Fin}$ expression halts and will only execute
when it is aborted by a parallel trail and is expanded by the~$\clear$
function.
%In Section~\ref{sec.formal.fins}, we show how to map a finalization block in
%the concrete language to a $fin$ in the formal semantics.
%
Note also that there is a syntactic restriction that postulates that~$\ceu{\Fin}$ bodies cannot
contain awaiting expressions ($\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$,
$\ceu{\Every}$, or $\ceu{\Fin}$),
i.e., the result of a~$\clear$ call is guaranteed to execute entirely within a reaction.

>>>>

\def\JOT{.8\jot}

@@

In this section, we introduce a reduced, abstract syntax for \CEU and
present an operational semantics that formalizes the behavior of its
programs.
% We describe a small synchronous kernel highlighting the peculiarities of \CEU,
% in particular, the stack-based execution for internal events.
%
The semantics deals only with the control aspects of \CEU.
Side-effects and system calls are encapsulated in a memory access
primitive and are assumed to behave like in conventional imperative
languages.

@@

The grammar below defines the syntax of a subset of \CEU that is
sufficient to describe all peculiarities of the language.
\bgroup
\def\lbl#1{\enspace\text{\emph{#1}}}%
\newdimen\X
\X=-1.3\jot
\begin{alignat*}{2}
  P\Coloneqq
      &\enspace\ceu{\Mem(\Id)}
      &&\lbl{any memory access to~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\AwaitExt(\Id)}
      &&\lbl{await external event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\AwaitInt(\Id)}
      &&\lbl{await internal event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\EmitInt(\Id)}
      &&\lbl{emit internal event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\Break}
      &&\lbl{loop escape}\\[\X]
      %%
  \mid&\enspace\ceu{\IfElse{\Mem(\Id)}{P_1}{P_2}}
      &&\lbl{conditional}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\,;\,P_2}
      &&\lbl{sequence}\\[\X]
      %%
  \mid&\enspace\ceu{\Loop P_1}
      &&\lbl{repetition}\\[\X]
      %%
  \mid&\enspace\ceu{\Every{\Id}\ P_1}
      &&\lbl{event iteration}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\And P_2}
      &&\lbl{par/and}\\[\X]
      %%
  \mid&\enspace\ceu{P_2\Or P_2}
      &&\lbl{par/or}\\[\X]
      %%
  \mid&\enspace\ceu{\Fin P}
      &&\lbl{finalization}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\AtLoop P_2}
      &&\lbl{unwinded loop}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\AtAnd\ P_2}
      &&\lbl{unwinded par/and}\\[\X]
      %%
  \mid&\enspace\ceu{P_1\AtOr\ P_2}
      &&\lbl{unwinded par/or}\\[\X]
      %%
  \mid&\enspace\ceu{\CanRun(n)}
      &&\lbl{can run on stack level~$n$}\\[\X]
      %%
  \mid&\enspace\ceu{\Nop}
      &&\lbl{terminated program}
\end{alignat*}
\egroup

The~$\ceu{\Mem(id)}$ primitive represents an assignment or system
call that affects the memory location identified by~$id$.\footnotemark\
%
Following the synchronous hypothesis, $\ceu{\Mem}$ statements and
expressions are considered to be atomic and instantaneous.
%
% As the challenging parts of \CEU reside on its control structures, we are not
% concerned here with a precise semantics for side effects, but only with their
% occurrences in programs.
%
%The special notation $nop$ is used to represent an innocuous $mem$ expression
%(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an
%unused identifier).
%
\footnotetext{Although the same symbol~$\ceu{\Id}$ is used in their
  definition, $\ceu{\Mem}$, $\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$
  and~$\ceu{\EmitInt}$ do not share identifiers: any identifier is either a
  variable, an external event, or an internal event.}

Most statements in the abstract language are mapped to their counterparts in
the concrete language.  The exceptions are the finalization
block~$\ceu{\Fin{p}}$ and the \texttt{@}-statements which do not exist in
the concrete language.  (These result from the expansion of the transition
rules to be presented.)

A further difference between the concrete and abstract languages regards the
emit-await pair.  In the concrete language, the \code{await} can be used as
an expression which evaluates to the value stored in the emitted
event, e.g., an
``\code{emit a(10)}'' awakes a ``\code{v=await a}'' setting
variable~\code{v} to~10.  Although the abstract $\ceu{\AwaitInt}$
and~$\ceu{\EmitInt}$ do not support such communication of values, it can be
easily simulated: one can use a shared variable to hold the value of an
$\ceu{\EmitInt}$ and access it after the corresponding $\ceu{\AwaitInt}$
awakes.

Finally, a ``\code{finalize $A$ with $B$ end; $C$}'' in the concrete
language is equivalent to ``\ceu{A;\;((\Fin{B})\ \Or\ C)}'' in the abstract
language.  In the concrete language, $A$ and~$C$ execute in sequence, and
the finalization code~$B$ is implicitly suspended waiting for~$C$
to terminate.  In the abstract language, ``$\ceu{\Fin B}$'' suspends forever
when reached (it is an awaiting statement that never awakes).  Hence, we
need an explicit \code{or} to execute~$C$ in parallel, whose termination
aborts ``$\ceu{\Fin B}$'', which finally causes~$B$ to execute (by the
semantic rules below).

@@

The operational semantics is a mathematical model that describes how an
abstract \CEU\ programs reacts to a single external input event, i.e., how
starting from this input event the program advances its state until all its
trails are blocked waiting for a another input event.
%%
The formalism we use here is that of small-step operational
semantics~\cite{Plotkin-G-D-1981} in which the meaning of programs is
defined in terms of transitions of an abstract machine.  Each transition
transforms a triple consisting of a program~$p$, a stack level~$n$, and an
emitted event~$e$ into a modified triple as follows:
\[
  \<p,n,e>\trans\<p',n',e'>\,,
\]
where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are non-negative
integers representing the current stack level, and~$e,e'\in\E\cup\{\nil\}$ are
the events emitted before and after the transition (both possibly being the
empty event~$\nil$).
%$\E$ is a set of primitive events.

We will refer to the triples on the left-hand and right-hand sides of
symbol~$\trans$ as \emph{descriptions} (denoted~$\delta$).  The description
on the left of~$\trans$ is called the \emph{input description},
the one on its right is called the \emph{output description}.

@@

At the beginning of a reaction to an input event~$id$, the input description is
initialized with stack level~0 ($n=0$) and with the externally emitted event~$e=id$.
%, but \code{emitInt} expressions can increase the stack level.
At the end of a reaction, after an arbitrary but finite number of transitions,
the last output description will block with a possibly modified program~$p'$ at stack
level~0 and with no event emitted:
\[
  \<p,0,e>\mathbin{\trans[*]}\<p',0,\nil>\,.
\]

We now proceed to give the rules for possible the transitions.  We
distinguish between two types of transitions:  \emph{outermost transitions}
$\out$ and \emph{nested transitions} $\nst$\,.

@@

The rules \R{push} and \R{pop} for~$\out$ transitions are non-recursive
definitions that apply to the program as a whole.  These are the only rules
that manipulate the stack level.
\begin{align*}
  &\AxiomC{$e\ne\nil$}
  \UnaryInfC{$\<p,n,e>\out\<\bcast(p,e),n+1,\nil>$}
  \DisplayProof
  \Rtag{push}\\[1.2\jot]
  %%
  &\AxiomC{$n>0$}
  \AxiomC{$p=\ceu{\Nop} \vee p=\ceu{\Break} \vee \isblocked(p,n)$}
  \BinaryInfC{$\<p,n,\nil>\out\<p,n-1,\nil>$}
  \DisplayProof
  \Rtag{pop}
\end{align*}

@@

Rule \R{push} can be applied whenever there is a nonempty event in the input
description,
and instantly broadcasts the event to the program, which means
    (a)~awaking any active $\ceu{\AwaitExt}$ or $\ceu{\AwaitInt}$ statements
    (see $\bcast$ in
        Figure~\ref{fig.bcast}),
    (b)~creating a nested reaction by increasing the stack level, and, at the same time,
    (c)~consuming the event ($e$ becomes~$\nil$).
%
Rule \R{push} is the only rule that matches an
emitted event and also immediately consumes it.

Rule \R{pop} decreases the stack level by one and can only be applied if the
program is blocked (see $\isblocked$
in Figure~\ref{fig.isblocked}) or terminated ($p=\ceu{\Nop}$ or $p=\ceu{\Break}$).
This condition ensures that an $\ceu{\EmitInt}$ only resumes after its internal
reaction completes and blocks in the current stack level.

At the beginning of a reaction, an external event is emitted, which
triggers rule \R{push}, immediately raising the stack level
to~1.
At the end of the reaction, the program will block or terminate and
successive applications of
rule~\R{pop} will lead to a description with this
same program at stack level~0.

@@

A complete reaction consists of a series of transitions of the form
\begin{align*}
  \<p,0,e_\ext>\outpush\<p_1,1,\nil>
  \Big[\null\nst[*]\null\out\null\Big]\!\!\ast
  \null\nst[*]\null\outpop\<p',0,\nil>\,,
\end{align*}
%
%-
% \begin{align*}
% a) &\5\5
%     \LL p,0,ext \RR
%         \1\xrightarrow[out]{push}\1
%     \LL q,1,\epsilon \RR
% \\
% b) &\5\5 \1[ \1\xrightarrow[in]{*}\1
%     \LL r,i,e \RR
%         \1\xrightarrow[out]\1
%     \LL s,j,\epsilon \RR \1]*
% \\
% c) &\5\5 \1\xrightarrow[in]{*}\1
%     \LL t,k,\epsilon \RR
%         \1\xrightarrow[out]{pop}\1
%     \LL u,0,\epsilon \RR
% \end{align*}
%-
%
where the~$\ast$'s stand for a finite (possibly zero) number of repetitions.
First, a~$\outpush$ starts a nested reaction at level~1.
Then, a series of alternations between zero or more~$\nst$ transitions (nested reactions) and a
single~$\out$ transition (stack operation) takes place.
Finally, a last~$\outpop$ transition decrements the
stack level to~0 and terminates the reaction.

@@

A $\ceu{\Mem}$ operation becomes a $\ceu{\Nop}$ which indicates the memory
access (rule \R{mem}).
An $\ceu{\EmitInt(id)}$ generates an event $\ceu{\Id}$ and becomes a
$\ceu{\CanRun(n)}$ which can only resume at level~$n$ (rule \R{emit-int}).
Since all~$\nst$ rules can only be applied if $e=\nil$, an $\ceu{\EmitInt}$
inevitably causes rule \R{push} to execute at the outer level, creating a new
level~$n+1$ on the stack.
Also, with the new stack level, the resulting $\ceu{\CanRun}(n)$ itself cannot
transition yet (rule~\R{can-run}), providing the desired stack-based semantics for
internal events.

The rules for conditionals and sequences are the following:
\vskip-.6\baselineskip
\begingroup
\def\JOT{.5\jot}
\begin{gather*}
  \AxiomC{$\eval(\ceu{\Mem(\Id)})$}
  \UnaryInfC{$\<\ceu{\IfElse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<p,n,\nil>$}
  \DisplayProof
  \Rtag{if-true}\\[\JOT]
  %%
  \AxiomC{$\lnot\eval(\ceu{\Mem(\Id)})$}
  \UnaryInfC{$\<\ceu{\IfElse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<q,n,\nil>$}
  \DisplayProof
  \Rtag{if-false}\\[\JOT]
  %%
  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
  \UnaryInfC{$\<\ceu{p\,;\,q},n,\nil>\nst\<\ceu{p';\,q},n,e>$}
  \DisplayProof
  \Rtag{seq-adv}\\[\JOT]
  \<\ceu{\Nop;\,q},n,\nil>\nst\<q,n,\nil>\Rtag{seq-nop}\\[\JOT]
  %%
  \<\ceu{\Break;\,q},n,\nil>\nst\<\ceu{\Break},n,\nil>\Rtag{seq-brk}
\end{gather*}
\endgroup
% \begin{align*}
%   \<\ceu{\Nop;\,q},n,\nil>&\nst\<q,n,\nil>\Rtag{seq-nop}\\[\JOT]
%   %%
%   \<\ceu{\Break;\,q},n,\nil>&\nst\<\ceu{\Break},n,\nil>\Rtag{seq-brk}
% \end{align*}

@@

Given that our semantics focuses on control, rules \R{if-true} and
\R{if-false} are the only to query~$\ceu{\Mem}$ as expressions.
%
Function~$\eval$ evaluates a~$\ceu{\Mem}$ expression to a boolean value.
%
%Although the value here is arbitrary, it is unique in a reaction, because a
%given expression can execute only once within it (remember that $loops$ must
%contain $awaits$ which, from rule \textbf{await}, cannot awake in the same
%reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).

@@

The rules for~$\ceu{\And}$ and~$\ceu{\Or}$ compositions ensure that their
left branch always transition before their right
branch:
%%
\begin{gather*}
  \hskip-.9em
  \<\ceu{p\And{q}},n,\nil>
  \nst\<\ceu{p\AtAnd(\CanRun(n);q)},n,\nil>
  \Rtag{and-expd}\\[\JOT]
  %%
  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
  \UnaryInfC{$\<\ceu{p\AtAnd{q}},n,\nil>\nst\<\ceu{p'\AtAnd{q}},n,e>$}
  \DisplayProof
  \Rtag{and-adv1}\\[\JOT]
  %%
  \AxiomC{$\isblocked(p,n)$}
  \AxiomC{$\<q,n,\nil>\nst\<q',n,e>$}
  \BinaryInfC{$\<\ceu{p\AtAnd{q}},n,\nil>\nst\<\ceu{p\AtAnd{q'}},n,e>$}
  \DisplayProof
  \Rtag{and-adv2}\\[1.2\jot]
% \end{gather*}
% \begin{gather*}
  \<\ceu{p\Or{q}},n,\nil>
  \nst\<\ceu{p\AtOr(\CanRun(n);q)},n,\nil>
  \Rtag{or-expd}\\[\JOT]
  %%
  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
  \UnaryInfC{$\<\ceu{p\AtOr{q}},n,\nil>\nst\<\ceu{p'\AtOr{q}},n,e>$}
  \DisplayProof
  \Rtag{or-adv1}\\[\JOT]
  %%
  \AxiomC{$\isblocked(p,n)$}
  \AxiomC{$\<q,n,\nil>\nst\<q',n,e>$}
  \BinaryInfC{$\<\ceu{p\AtOr{q}},n,\nil>\nst\<\ceu{p\AtOr{q'}},n,e>$}
  \DisplayProof
  \Rtag{or-adv2}
\end{gather*}

@@

Rules~\R{and-expd} and~\R{or-expd} insert a~$\ceu{\CanRun(n)}$ at the beginning
of the right branch.
This ensures that~any $\ceu{\EmitInt}$ on the left branch, which eventually becomes
a~$\ceu{\CanRun(n)}$, resumes before the right branch starts.
%
The deterministic behavior of the semantics relies on the \emph{isblocked}
predicate (see Figure~\ref{fig.isblocked}) which is used in rules
\R{and-adv2} and \R{or-adv2}.
These rules require the left branch~$p$ to be blocked for the
right branch to transition from~$q$ to~$q'$.

In a parallel~$\ceu{\AtAnd}$, if one branch terminates, the composition becomes the other branch (rules \R{and-nop1} and
\R{and-nop2} below).
%
In a parallel~$\ceu{\AtOr}$, however, if one branch terminates, the
whole composition
terminates and~$\clear$ is called to finalize the aborted
branch (rules \R{or-nop1} and \R{or-nop2}).

\begingroup
\begin{gather*}
  \<\ceu{{\Nop}\AtAnd{q}},n,\nil>\nst\<q,n,\nil>\Rtag{and-nop1}\\[\JOT]
  %%
  \AxiomC{$\isblocked(p,n)$}
  \UnaryInfC{$\<\ceu{p\AtAnd{\Nop}},n,\nil>\nst\<p,n,\nil>$}
  \DisplayProof
  \Rtag{and-nop2}\\[\JOT]
  %%
  \<\ceu{{\Nop}\AtOr{q}},n,\nil>\nst\<\clear(q),n,\nil>\Rtag{or-nop1}\\[\JOT]
  %%
  \AxiomC{$\isblocked(p,n)$}
  \UnaryInfC{$\<\ceu{p\AtOr{\Nop}},n,\nil>\nst\<\clear(p),n,\nil>$}
  \DisplayProof
  \Rtag{or-nop2}
\end{gather*}
\endgroup

@@

The~$\clear$ function (see Figure~\ref{fig.clear}) concatenates all
active~$\ceu{\Fin}$ bodies of the branch being aborted, so that they execute before the
composition rejoins.

As there are no transition rules for~$\ceu{\Fin}$ statements,
 once reached, a $\ceu{\Fin}$ halts and will only be consumed
if its trail is aborted.  At this point, its body will
execute as a result of the~$\clear$ call.  The body of a~$\ceu{\Fin}$
statement always execute withing a reaction.  This is due to a syntactic
restriction: $\ceu{\Fin}$ bodies cannot
contain awaiting statements (namely, $\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$,
$\ceu{\Every}$, or $\ceu{\Fin}$).
